---
title: |
  | KU Leuven Summer School   
  | Segment 2C  
  | Missing (Perhaps?) Not at Random 
author: Paul Gustafson
date: September 15, 2022
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenrs.tex
---

```{r, echo=F}
### global reproducibility
set.seed(1234)
### sane output
options(digits=3)
options(width=65)
```

# Minimal Working Example

```{r}
expit <- function(z) {1/(1+exp(-z))}
logit <- function(p) {log(p)-log(1-p)}

n <- 2000
x <- rbinom(n, size=1, prob=.4)
y <-rbinom(n, size=1, 
             prob=expit(cbind(1,x) %*% 
                        c(-1, 0.2)))
```

# Creating the problem

```{r}
r <- rbinom(n, size=1,
  prob=expit(0.1 + 0.3*x - 0.6*x*y))

x.obs <- rep(NA, n); x.obs[r==1] <- x[r==1]
dat <- data.frame(x=x.obs, y=y)
```

# MAR analysis (i)

```{r}
genmod.mar.string <- "
model {
  alpha ~ dnorm(0, 0.1)
  beta0 ~ dnorm(0, 0.1)
  beta1 ~ dnorm(0, 0.1)

  for (i in 1:n) {
    x[i] ~ dbern(pr.x)

    y[i] ~ dbern(pr.y[i])
    logit(pr.y[i]) <- beta0 + beta1*x[i]
  }
  logit(pr.x) <- alpha
}"
```
# MAR analysis (ii)

```{r, echo=F, message=F, warning=F}
require(rjags)
require(MCMCvis)
```

```{r, message=F, warning=F, results="hide", cache=T}
### generative model, data go in
mod <- jags.model(
  textConnection(genmod.mar.string),
  data=list(x=dat$x, y=dat$y,
            n=dim(dat)[1]), 
         n.chains=3)

update(mod, 2000) # burn-in

###  MC output comes out
opt.mar.JAGS <- coda.samples(mod, n.iter=15000,
  variable.names=c("alpha","beta0","beta1")) 
```
# MAR Analysis (iii)

```{r}
MCMCsummary(opt.mar.JAGS)
```

# Now explictly model the missingness 

```{r}
statmod.mnar.string <- "
for (i in 1:n) {
  x[i] ~ dbern(pr.x)

  y[i] ~ dbern(pr.y[i])
  logit(pr.y[i]) <- beta0 + beta1*x[i]
  
  r[i] ~ dbern(pr.r[i])
  logit(pr.r[i]) <- gamma0 + gamma1*y[i] +
    gamma2*x[i] + (gamma3-gamma2)*x[i]*y[i]
}
logit(pr.x) <- alpha"
```

# With this prior specification

```{r}
prior.mnarA.string <- "
   alpha ~ dnorm(0, 0.1)
   beta0 ~ dnorm(0, 0.1)
   beta1 ~ dnorm(0, 0.1)
  gamma0 ~ dnorm(0, 0.1)
  gamma1 ~ dnorm(0, 0.1)
  gamma2 ~ dnorm(0, 100)
  gamma3 ~ dnorm(0, 100)"
```

```{r}
genmod.mnarA.string <-  paste(
  "model {",prior.mnarA.string, 
  statmod.mnar.string, "}")
```

# Any pertinent remarks about our generative model?

# Turn the crank

```{r, message=F, warning=F, results="hide", cache=T}
### generative model, data go in
mod <-  jags.model(
  textConnection(genmod.mnarA.string),
  data=list(x=dat$x, y=dat$y,
            r=as.vector(!is.na(dat$x)),
            n=dim(dat)[1]),
           n.chains=3)

update(mod,2000) # burn-in

###  MC output comes out
opt.mnarA.JAGS <- coda.samples(
  mod,
  n.iter=15000,
  variable.names=c("alpha","beta0","beta1",
  "gamma0","gamma1","gamma2","gamma3")) 
```
# And report posterior quantities

```{r}
MCMCsummary(opt.mnarA.JAGS)
```

# And now the same with this prior specification

```{r}
prior.mnarB.string <- "
   alpha ~ dnorm(0, 0.1)
   beta0 ~ dnorm(0, 0.1)
   beta1 ~ dnorm(0, 0.1)
  gamma0 ~ dnorm(0, 0.1)
  gamma1 ~ dnorm(0, 0.1)
  gamma2 ~ dnorm(0, 25)
  gamma3 ~ dnorm(0, 25)"
```

```{r}
genmod.mnarB.string <-  paste(
  "model {",prior.mnarB.string, 
  statmod.mnar.string, "}")
```

# Crank

```{r, message=F, warning=F, results="hide", cache=T}
### generative model, data go in
mod <-  jags.model(
  textConnection(genmod.mnarB.string),
  data=list(x=dat$x, y=dat$y,
    r=as.vector(!is.na(dat$x)),
    n=dim(dat)[1]),
  n.chains=3)

update(mod, 2000) # burn-in

###  MC output comes out
opt.mnarB.JAGS <- coda.samples(
  mod,
  n.iter=15000,
  variable.names=c("alpha","beta0","beta1","gamma0",
    "gamma1","gamma2","gamma3")) 
```

# Answer

```{r}
MCMCsummary(opt.mnarB.JAGS)
```

# Take a bit of stock

Recall $\gamma_2$, $\gamma_{3}$ describe $\log OR(R,X|Y=y)$

\begin{center}
\begin{tabular}{ccc}
           &               & quality of  \\
prior SDs  & posterior SDs & numerical approximation \\
\end{tabular}
\end{center}

\vspace{2.in}

# Adjective soup

\begin{center}
\begin{tabular}{cc}
MAR                                    & MNAR\\
$\gamma_2=\gamma_3=0$ known   &         $\gamma_{2},\gamma_3$ unknown\\
(and don't need to model $R$) &  \\
\end{tabular}
\end{center}

\vspace{2.in}

# Folk theorem


# Important distinction related to folk theorem











