---
title: |
  | KU Leuven Summer School   
  | Segment 3B
  | More Misclassification Models 
author: Paul Gustafson
date: September 15, 2022 
output: 
  beamer_presentation:
    keep_tex: true
    includes:
      in_header: header_pagenrs.tex
urlcolor: blue
---

```{r, echo=F}
### global reproducibility
set.seed(1234)
### sane output
options(digits=3)
options(width=65)
```

```{r, echo=F, message=F, warning=F}
require(rjags)
require(MCMCvis)
```

```{r, echo=F}
logit <- function(p) { log(p)-log(1-p) }
expit <- function(z) { 1/(1+exp(-z)) }
```



# Start with a mystery dataset

Case-control again, but:

* Nobody has a measurement of true exposure $X$

* Everbody has a pair of measurements from two **different** surrogates (for $X$), $X^{*}_{1}$, $X^{*}_{2}$

E.g., think $X_1^{*}$ $\sim$ self-report, $X^{*}_{2}$ $\sim$ imperfect lab test

```{r, echo=F}
### mystery data
n <- 5000

### balanced case-control study
y <- sample(c(rep(0, n/2), rep(1, n/2)))

### true OR of 2
gamma.0.TR <- 0.1
gamma.1.TR <- expit(logit(gamma.0.TR)+log(2))

x <- rbinom(n, size=1, prob=(1-y)*gamma.0.TR + y*gamma.1.TR)

### two independent surrogates
xstr1 <- rbinom(n, size=1, prob=(1-x)*(1-0.85)+x*(0.75))
xstr2 <- rbinom(n, size=1, prob=(1-x)*(1-0.95)+x*(0.60))

### Mother Nature WITHHOLDS the actual x
rm(x)

dta <- data.frame(xstr1=xstr1, xstr2=xstr2, y=y)
```

```{r}
head(dta)
```
# Mystery dataset, continued 

```{r}
table(dta)
```

# Some simple analyses


```{r}
summary(glm(y~xstr1, family=binomial))$coef
summary(glm(y~xstr2, family=binomial))$coef
```

# Generative model

$f(\mbox{params})$
$\prod_{i=1}^{n}$
${\color{gray}f(y_i)}$
$f(x_i|y_i)$
$f(x^{*}_{1,i}, x^{*}_{2,i}|x_i,{\color{gray}y_i})$

\vspace{2.in}


# This will run (but ess/wall-time unpleasant)

```{r}
genmod.string <- "model{
  
### prior distribution
gamma.0 ~ dunif(0,1)
gamma.1 ~ dunif(0,1)
sn1 ~ dunif(0.5, 1)
sp1 ~ dunif(0.5, 1)
sn2 ~ dunif(0.5, 1)
sp2 ~ dunif(0.5, 1)

trgt <- logit(gamma.1)-logit(gamma.0)

for (i in 1:n) {
  x[i] ~ dbern((1-y[i])*gamma.0+y[i]*gamma.1)
  xstr1[i] ~ dbern((1-x[i])*(1-sp1)+x[i]*sn1)
  xstr2[i] ~ dbern((1-x[i])*(1-sp2)+x[i]*sn2)
}  

}" 
```

# Instead consider

```{r}
genmod.string <- "model {
gamma.0 ~ dunif(0,1); gamma.1 ~ dunif(0,1)
trg <- logit(gamma.1)-logit(gamma.0)
sn1 ~ dunif(0.5,1); sp1 ~ dunif(0.5,1)
sn2 ~ dunif(0.5,1); sp2 ~ dunif(0.5,1)

### controls: dist(xstr1, xstr2 |y=0)
q.0[1] <- (1-gamma.0)*sp1*sp2 + gamma.0*(1-sn1)*(1-sn2)
q.0[2] <- (1-gamma.0)*(1-sp1)*sp2 + gamma.0*(sn1)*(1-sn2)
q.0[3] <- (1-gamma.0)*sp1*(1-sp2) + gamma.0*(1-sn1)*sn2
q.0[4] <- (1-gamma.0)*(1-sp1)*(1-sp2) + gamma.0*sn1*sn2
dat.0 ~ dmulti(q.0[], n.0)

### cases: dist(xstr1, xstr2 |y=1)
q.1[1] <- (1-gamma.1)*sp1*sp2 + gamma.1*(1-sn1)*(1-sn2)
q.1[2] <- (1-gamma.1)*(1-sp1)*sp2 + gamma.1*(sn1)*(1-sn2)
q.1[3] <- (1-gamma.1)*sp1*(1-sp2) + gamma.1*(1-sn1)*sn2
q.1[4] <- (1-gamma.1)*(1-sp1)*(1-sp2) + gamma.1*sn1*sn2
dat.1 ~ dmulti(q.1[], n.1)
}"
```

# Pause: what's going on here?

# Pause, continued

# Turn the crank

```{r, cache=F,message=F, warning=F, results='hide'}
### generative model, data go in
mod <- jags.model(textConnection(genmod.string),
  data=list(dat.0=as.vector(table(dta)[,,1]), 
            n.0=sum(table(dta)[,,1]),
            dat.1=as.vector(table(dta)[,,2]), 
            n.1=sum(table(dta)[,,2])),
  n.chains=3)

update(mod,2000) #burn-in

### MCMC output comes out
opt.JAGS <- coda.samples(mod, n.iter=60000, thin=10, 
  variable.names=c("gamma.0","gamma.1","sn1","sp1",
                   "sn2","sp2","trg"))
```

# Inference

```{r}
MCMCsummary(opt.JAGS)
```

# Inference, continued

```{r}
MCMCplot(opt.JAGS)
```

# And a grand reveal

# Pause to marvel for a moment: Asked a lot of these data, and they delivered

# An aside on parameter-counting (1 of 2)

\vspace{3.in}


Relevant ref: [Hui \& Walter (1980, Biom.)](https://doi.org/10.2307/2530508)



# An aside on parameter-counting (2 of 2)

\vspace{3.in}

Relevant ref: [Johnson \& Hanson (2005, Stat. Sci., comment)](https://doi.org/10.1214/088342305000000098)

# Pause some more: Lunch is never *completely* free 

# And yet another sense in which lunch isn't free




